Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-04-02 application.tex
first complete revision after barker & latter

\section{Application to protoplanetary disks}\label{application} 
As an application of our linear models, we estimate where in a
protoplanetary disk do the thermodynamic conditions allow the VSI to
operate. For a realistic disk one can consider the energy equation
with radiative cooling, 
\begin{align}\label{real_energy}
  % \rho T \frac{DS}{Dt} = - \left(\nabla\cdot\bm{F} - Q_+\right), 
  \frac{\p P}{\p t} + \bm{v}\cdot\nabla P +\gamma P \nabla\cdot\bm{v} = - \frac{P}{\rho T
    c_v}\left(\nabla\cdot\bm{F} - Q_+\right), 
\end{align}
where $c_v$ is the heat capacity at constant
volume, 
$\bm{F}$ is the radiative flux and $Q_+$ is a heating
term to permit an equilibrium solution.

@@

\subsection{A model for combined thermal relaxation}\label{toy_relax}
%To unify $t_\mathrm{cool}$ and $t_\mathrm{diff}$, 
We define the effective thermal relaxation timescale in linear theory as
\begin{align}\label{tc_def}
  t_c &\equiv t_\mathrm{cool} + t _\mathrm{diff}. %=
  % \frac{l_\mathrm{rad}^2}{3\eta} + \frac{1}{\eta k_x^2},  
\end{align}
Eq. \ref{tc_def} is a simple prescription so
that for small scales (large $k_x$), $t_c\to t_\mathrm{cool}$, while
for large scales (small $k_x$), $t_c\to t_\mathrm{diff}$. Writing the
volume density $\rho = \Sigma\hat{\rho}/\sqrt{2\pi}H$, where
$\hat{\rho} = \rho/\rho_0$ and $\Sigma$ is the surface density at the
radius of interest, the dimensionless thermal
relaxation time $\beta$ becomes 
\begin{align}\label{real_beta}
  \beta(z;\khat) \equiv t_c\Omega_k =
  % \frac{1}{\hat{\eta}}\left[\frac{l_\mathrm{rad}^2}{3H^2}
  %   + \frac{\hat{\rho}^2(z)}{2\pi\khat^2}\right]  \simeq
  \frac{\Sigma^2\Omega_k}{\eta\rho^2}\left[\frac{1}{3\kappa_d^2\Sigma^2} 
    + \frac{\hat{\rho}^2(z)}{2\pi \khat^2}\right].
\end{align}
Note that $\beta$ depends on both the perturbation lengthscale $\khat$ 
and distance from the midplane.

\subsection{Evaluation for a fiducial disk model}
We evaluate Eq. \ref{real_beta} for the protoplanetary disk model
described in \cite{chiang10} and summarized in Appendix \ref{mmsn}
(where an opacity model is also chosen). Defining $r_\mathrm{AU}\equiv 
r/\mathrm{AU}$, we find Eq. \ref{real_beta} becomes
\begin{align}\label{beta_mmsn}
  \beta(z; \khat) =
  &\frac{4.4\times10^5}{\mu(\gamma-1)}\left(\frac{\hat{\kappa}_d\hat{\Sigma}^2}{\hat{T}}\right)r_\mathrm{AU}^{-57/14}\notag\\ 
&\times\left[\frac{8.3\times10^{-9}}{\hat{\kappa}_d^2\hat{T}^4\hat{\Sigma}^2}r_\mathrm{AU}^{33/7}+\frac{\hat{\rho}^2(z)}{2\pi\khat^2}\right].         
\end{align} 
We refer to the case with $\gamma=1.4$, $\mu=2.33$ and the scalings
$\hat{\Sigma}=\hat{\kappa}_d=\hat{T}=1$ as the Minimum Mass Solar
Nebulae (MMSN) and $\beta=\beta_\mathrm{MMSN}$, 
\begin{align}\label{beta_mmsn_simp}
  \beta_\mathrm{MMSN} = 3.9\times10^{-3} r_\mathrm{AU}^{9/14}\left( 1 +
    1.9\times10^7\hat{\rho}^2r_\mathrm{AU}^{-33/7}\khat^{-2}\right). 
\end{align}
% where
% \begin{align}
%   \
% %  \frac{3.7\times10^{-3}}{\mu\left(\gamma-1\right)\hat{\kappa}_d\hat{T}^5}, 
% \end{align}
% and
% \begin{align}
%   \khat_0 = 4.4\times10^{3}\hat{\kappa}_d\hat{T}^2\hat{\Sigma}.
% \end{align}
It is instructive to compare the above thermal timescale 
with the criterion Eq. \ref{iso_vsi_cond} evaluated for the MMSN,     
\begin{align}\label{bcrit_mmsn}
  % \beta_\mathrm{crit} = \frac{1.44\times10^{-2}}{(\gamma
  % -1)}\left(\frac{\hat{T}}{\mu}\right)^{1/2}r_\mathrm{AU}^{2/7}. 
  \beta_\mathrm{crit,MMSN} = 0.024r_\mathrm{AU}^{2/7}. 
\end{align}

@@

In the inner few AU of the MMSN, however, $\beta$ 
may be smaller or greater than $\beta_\mathrm{crit}$ depending on
$z$ as the disk transitions from an optically thick midplane to an
optically thin atmosphere. 
%Given a fixed $\khat$, one can have $\beta > \beta_\mathrm{crit}$ at
%the optically thick midplane, but with increasing height the disk 
%becomes optically thin and $\beta<\beta_\mathrm{crit}$. 
Our analytical criteria is not valid for non-uniform $\beta$, and we must
turn to numerical solutions of the linear problem to determine whether
or not the VSI operates. We do this in the next section.

@@

We remark that it is possible to satisfy $\beta < \beta_\mathrm{crit}$ 
for sufficiently large $\khat$. We illustrate this in 
Fig. \ref{mmsn_bcrit_bcool_mink} by plotting at each radius the value
of $\khat$ such that $\beta = \beta_\mathrm{crit}$, which show that in
the inner disk, say $r\sim 1$AU, the VSI would only occur on very
small radial lengthscales. 
%surface modes can develop for large $\khat$, 
%which have small vertical lengthscales scales. 
%also small vertical scales if they are surface modes

@@

In Fig. \ref{mmsn_overall} we plot characteristic growth timescales
$t_\mathrm{grow} \equiv 1/\nu$ of the fundamental mode with a range of
wavenumbers, in units of the local orbital time $P_\mathrm{orb}\equiv
2\pi/\Omega_k$. In fact, we either only find the fundamental mode, or 
that it is the most unstable, except for the appearance of some 
surface modes towards the inner disk when considering large 
wavenumbers $\khat\gtrsim 50$. However, surface modes are dependent on
boundary conditions in a vertically isothermal disk \citep{barker15},  
and disturbances with large $\khat$ are more likely subject to viscous
decay, as discussed above. % It is also unclear if surface modes are
% dynamically important given they are concentrated in the disk
% atmosphere where the density is small. 
% The dominance of the fundamental mode is also
% observed in hydrodynamic simulations of \cite{stoll14} which treat
% radiative cooling properly.   unclear if surface modes are
% dynamically important because the there is little mass in the disk at

We therefore consider Fig. \ref{mmsn_overall} as a representation of the
VSI in the MMSN. We see the VSI operates from $\sim 5$AU to $\sim
50$AU with growth timescales $\sim 30$---40 orbits and
$t_\mathrm{grow}$ rises rapidly towards the inner disk. It is
necessary to consider smaller scales with decreasing $r$ as the disk
generally becomes optically thick.

@@

We thus find Fig. \ref{mmsn_overall} to be roughly consistent with
comparing the midplane thermal relaxation timescales to the critical
thermal timescale. For example, 
Fig. \ref{mmsn_bcrit_bcool} show that $\beta(\khat=10,z=0) \gtrsim
\beta_\mathrm{crit}$ for $r\lesssim 10$AU in the MMSN (left middle panel); and Fig. \ref{mmsn_overall}
show that modes with $\khat=10$ are stabilized inside $\sim 8$AU (where
$t_\mathrm{grow}\to\infty$). Our numerical results suggest the VSI
is efficient at radii of few tens of AU in the MMSN, consistent with
estimates made by \cite{nelson13}.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/eigen_compare_grow.ps}
  \caption{Characteristic growth times of the VSI in 
    the MMSN. These are calculated using the scale and height
    dependent thermal relaxation given by Eq. \ref{beta_mmsn_simp}. 
    \label{mmsn_overall}}    
\end{figure}

>>>>

\section{Application to protoplanetary disks}\label{application} 
As an application of our linear models, we estimate where in a
protoplanetary disk do the thermodynamic conditions allow the VSI to
operate. For a realistic disk one might consider an energy equation
with radiative cooling, 
\begin{align}\label{real_energy}
  % \rho T \frac{DS}{Dt} = - \left(\nabla\cdot\bm{F} - Q_+\right), 
  \frac{\p P}{\p t} + \bm{v}\cdot\nabla P +\gamma P \nabla\cdot\bm{v} = - \frac{P}{\rho T
    c_v}\left(\nabla\cdot\bm{F} - Q_+\right), 
\end{align}
where $c_v$ is the heat capacity at constant
volume, 
$\bm{F}$ is the radiative flux and $Q_+$ is a heating
term to permit an equilibrium solution.

@@

\subsection{A model for combined thermal relaxation}\label{toy_relax}
%To unify $t_\mathrm{cool}$ and $t_\mathrm{diff}$, 
We define the effective thermal relaxation timescale in linear theory as
\begin{align}\label{tc_def}
  t_c &\equiv t_\mathrm{cool} + t _\mathrm{diff}. %=
  % \frac{l_\mathrm{rad}^2}{3\eta} + \frac{1}{\eta k_x^2},  
\end{align}
Eq. \ref{tc_def} is a simple prescription so
that for small scales (large $k_x$), $t_c\to t_\mathrm{cool}$, while
for large scales (small $k_x$), $t_c\to t_\mathrm{diff}$. Writing the
volume density $\rho = \Sigma\hat{\rho}/\sqrt{2\pi}H$, where
$\hat{\rho} = \rho/\rho_0$ and $\Sigma(r)$ is the surface density at the
radius of interest, the dimensionless thermal
relaxation time $\beta$ becomes 
\begin{align}\label{real_beta}
  \beta(r,z;\khat) \equiv t_c\Omega_k =
  % \frac{1}{\hat{\eta}}\left[\frac{l_\mathrm{rad}^2}{3H^2}
  %   + \frac{\hat{\rho}^2(z)}{2\pi\khat^2}\right]  \simeq
  \frac{\Sigma^2\Omega_k}{\eta\rho^2}\left[\frac{1}{3\kappa_d^2\Sigma^2} 
    + \frac{\hat{\rho}^2(z)}{2\pi \khat^2}\right].
\end{align}
Note that $\beta$ depends the perturbation lengthscale $\khat$,
distance from the midplane, as well as the distance from the star.

\subsection{Evaluation for a fiducial disk model}
We evaluate Eq. \ref{real_beta} for the protoplanetary disk model
described in \cite{chiang10} and summarized in Appendix \ref{mmsn}
(where an opacity model is also chosen). Defining $r_\mathrm{AU}\equiv 
r/\mathrm{AU}$, we find Eq. \ref{real_beta} becomes
\begin{align}\label{beta_mmsn}
  \beta(r, z; \khat) =
  &\frac{4.4\times10^5}{\mu(\gamma-1)}\left(\frac{\hat{\kappa}_d\hat{\Sigma}^2}{\hat{T}}\right)r_\mathrm{AU}^{-57/14}\notag\\ 
&\times\left[\frac{8.3\times10^{-9}}{\hat{\kappa}_d^2\hat{T}^4\hat{\Sigma}^2}r_\mathrm{AU}^{33/7}+\frac{\hat{\rho}^2(z)}{2\pi\khat^2}\right].         
\end{align} 
We refer to the case with $\gamma=1.4$, $\mu=2.33$ and the scalings
$\hat{\Sigma}=\hat{\kappa}_d=\hat{T}=1$ as the Minimum Mass Solar
Nebulae (MMSN) and $\beta=\beta_\mathrm{MMSN}$, 
\begin{align}\label{beta_mmsn_simp}
  \beta_\mathrm{MMSN} = 3.9\times10^{-3} r_\mathrm{AU}^{9/14}\left( 1 +
    1.9\times10^7\hat{\rho}^2r_\mathrm{AU}^{-33/7}\khat^{-2}\right). 
\end{align}
% where
% \begin{align}
%   \
% %  \frac{3.7\times10^{-3}}{\mu\left(\gamma-1\right)\hat{\kappa}_d\hat{T}^5}, 
% \end{align}
% and
% \begin{align}
%   \khat_0 = 4.4\times10^{3}\hat{\kappa}_d\hat{T}^2\hat{\Sigma}.
% \end{align}
It is instructive to compare the above thermal timescale 
with the criterion Eq. \ref{iso_vsi_cond} evaluated for the MMSN,     
\begin{align}\label{bcrit_mmsn}
  % \beta_\mathrm{crit} = \frac{1.44\times10^{-2}}{(\gamma
  % -1)}\left(\frac{\hat{T}}{\mu}\right)^{1/2}r_\mathrm{AU}^{2/7}. 
  \beta_\mathrm{crit,MMSN} = 0.024r_\mathrm{AU}^{2/7}. 
\end{align}

@@

In the inner few AU of the MMSN, however, $\beta$ 
may be smaller or greater than $\beta_\mathrm{crit}$ depending on
$z$ as the disk transitions from an optically thick midplane to an
optically thin atmosphere (see also Fig. \ref{bcrit_mmsn2d}). 
%Given a fixed $\khat$, one can have $\beta > \beta_\mathrm{crit}$ at
%the optically thick midplane, but with increasing height the disk 
%becomes optically thin and $\beta<\beta_\mathrm{crit}$. 
Our analytical criteria $\beta < \beta_\mathrm{crit}$ is not valid for
vertically non-uniform $\beta$, and we must turn to numerical
solutions of the linear problem to determine whether or not the VSI
operates. We do this in the next section.

@@

We remark that it is possible to satisfy $\beta < \beta_\mathrm{crit}$ 
for sufficiently large $\khat$. We illustrate this in 
Fig. \ref{mmsn_bcrit_bcool_mink} by plotting at each radius the value
of $\khat$ such that $\beta = \beta_\mathrm{crit}$. If we consider the
midplane of the inner disk, say $r\sim 1$AU, the VSI would only occur
on very small radial lengthscales $\ll O(10^{-2}H)$. In the disk
atmosphere where $\beta$ is small, surface modes may develop, 
%However, at such
%small scales, surface modes are likely to dominate (and are
%restricted to the disk atmosphere), 
but they are not well-defined in vertically isothermal disk models
with an imposed boundary, and they also require large $\khat$.

%surface modes can develop for large $\khat$, 
%which have small vertical lengthscales scales. 
%also small vertical scales if they are surface modes

@@

In Fig. \ref{mmsn_overall} we plot characteristic growth timescales
$t_\mathrm{grow} \equiv 1/\nu$ of the fundamental mode with a range of
wavenumbers, in units of the local orbital time $P_\mathrm{orb}\equiv
2\pi/\Omega_k$. In fact, we either only find the fundamental mode, or 
that it is the most unstable, except for the appearance of some 
surface modes for large wavenumbers $\khat\gtrsim 50$, which are 
boundary-dependent.  
%However, surface modes are dependent on
%boundary conditions in a vertically isothermal disk \citep{barker15},  
%and 
Disturbances with large $\khat$ are also more likely subject to viscous 
decay, as discussed above.

% It is also unclear if surface modes are
% dynamically important given they are concentrated in the disk
% atmosphere where the density is small. 
% The dominance of the fundamental mode is also
% observed in hydrodynamic simulations of \cite{stoll14} which treat
% radiative cooling properly.   unclear if surface modes are
% dynamically important because the there is little mass in the disk at

We therefore consider Fig. \ref{mmsn_overall} as a representation of the
VSI in outer parts of the MMSN. We see the VSI operates from $\sim 5$AU to $\sim
50$AU with growth timescales $\sim 30$---40 orbits and
$t_\mathrm{grow}$ rises rapidly towards the inner disk. It is
necessary to consider smaller scales with decreasing $r$ as the disk
generally becomes optically thick.

@@

We thus find Fig. \ref{mmsn_overall} to be roughly consistent with
comparing the midplane thermal relaxation timescales to the critical
thermal timescale. For example, 
Fig. \ref{mmsn_bcrit_bcool} show that $\beta(z=0,\khat=10) \gtrsim
\beta_\mathrm{crit}$ for $r\lesssim 10$AU in the MMSN (left middle
panel); and Fig. \ref{mmsn_overall} 
show that modes with $\khat=10$ are stabilized inside $\sim 8$AU (where
$t_\mathrm{grow}\to\infty$). Our numerical results suggest the VSI
is efficient at radii of few tens of AU in the MMSN, consistent with
estimates made by \cite{nelson13}.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/eigen_compare_grow.ps}
  \caption{Characteristic growth times of the VSI in 
    the MMSN. These are calculated using the thermal relaxation
    function given by Eq. \ref{beta_mmsn_simp}. 
    \label{mmsn_overall}}    
\end{figure}

### 2015-04-02 numerical.tex
first complete revision after barker & latter

The modes shown in Fig. \ref{compare_modes_iso_kx10} are `body modes'
and are inertial waves destabilized by vertical shear
\citep{barker15}. (The exception being the fastest growing
mode for $\zmax=7H$ which is a surface mode --- see the discussion
below.)  The `fundamental corrugation mode', marked by `f',
corresponds to the eigenvalue with the smallest $|\sigma|$. It is robust
against vertical boundary conditions as it appears at the same
position in Fig. \ref{compare_modes_iso_kx10} for all cases. Higher  
order modes, corresponding to increasing $|\omega|$, increasingly
deviates from the analytic line due to the imposed boundaries, 
unless $\zmax$ is increased \citep{barker15}.

@@

\subsubsection{Comment on `surface modes'}\label{surf_comment} 
Fig. \ref{compare_modes_iso_kx10} lack the distinct `surface modes' found by 
\cite{nelson13}, \cite{mcnally14}  and \cite{barker15}. 
This a class of unstable modes associated with vertical shear near the
disk boundaries \citep{barker15}. They are absent in our  
analysis because there we assumed a disk with infinite 
vertical extent, since the vertically isothermal disk has no
surface. However, surface modes can occur in the present calculations 
because there is an imposed vertical boundary.

We illustrate surface modes with the mode diagram for $\khat=30$ in
Fig. \ref{compare_modes_iso_kx30}. Surface 
modes lie almost parallel to the vertical axis
in such diagrams. They are readily identified for  $\zmax\geq 5H$ and
are marked by `s'.  Fig. \ref{lowfreq_eigenfunc_surf} shows an example
of a surface mode. We typically find an increasing
number of surface modes with increasing $\zmax$ and/or $\khat$,
consistent with \cite{barker15}.

@@

We find the number of unstable modes decrease with increasing
$\beta$. Consider the fiducial setup with $\zmax=5H$ 
(middle panel). The fundamental mode ($M=0$) is not the most unstable for   
$\beta=0.01$ as there are modes with $M>0$ that have larger growth 
rates. As remarked above, these higher order modes are affected by
$\zmax$, e.g. there is a significant mismatch between numerical and 
analytical results for $M\geq1$ when $\zmax=3H$ (upper panel).

@@

Fig. \ref{lowfreq_eigenfunc_cool}---\ref{lowfreq_eigenfunc_2d_cool}
shows the fundamental mode with $\khat=10$ in a disk with 
$\beta=0.1$. Comparing these plots with
Fig. \ref{lowfreq_eigenfunc}---\ref{lowfreq_eigenfunc_2d}, which have
$\beta=10^{-3}$, we see that increasing $\beta$ restricts the region 
characterized by $W\sim z$ and $\delta v_z\sim$ constant closer to the
midplane, with oscillatory behavior emerging near vertical boundaries
where buoyancy becomes important. This leads to the `tilted' pressure
perturbations seen in Fig. \ref{lowfreq_eigenfunc_2d_cool} 
(cf. Fig. \ref{lowfreq_eigenfunc_2d}).

@@

Note that the curves are not neccessarily smooth because the most unstable
mode may not always correspond to the same type of disturbance. For example,
as seen above, for $\khat \leq 10$ higher order body modes are
dominant as $\beta\to 0$, but the fundamental corrugation
mode becomes dominant as $\beta$ is increased. For $\khat=1$ the VSI
can operate slightly beyond $\beta_\mathrm{crit}$  (recall that
$\beta_\mathrm{crit}$ was derived assuming $\khat^2\gg 1$). For
$\khat=5,\,10$ (and $30$ for $\zmax=3H$), we find
$\beta_\mathrm{crit}$ accurately predicts the thermal timescal beyond
which the VSI is suppressed.

Complications from vertical boundaries arise when considering large
$\khat$. For $\khat=30$ and $\zmax=5H,\,7H$ we find for $\beta\gtrsim
0.1$ the dominant disturbance correspond to the leftmost eigenvalues 
seen in the lower two panels of Fig. \ref{compare_modes_cool_kx30},
which do not exist for $\zmax=3H$. For $\khat=50,\,100$ and
$\beta\to0$, we find surface modes dominate so the maximum growth rate
depends on $\zmax$. However, the maximum growth rate is insensitive to
$\zmax$ for moderately large $\beta$ ($\gtrsim 0.1$), i.e. close to
the stabilization of the fundamental mode. 
%not surprising since we enable strongly stabilizing buoyancy forces
%at large zmax

@@

We further  test the dependence of $\beta_\mathrm{crit}$ on disk parameters
(Eq. \ref{iso_vsi_cond}).  We use the fiducial setup as a  reference
point and vary $q\in[-0.2,-1.2]$,  $\gamma\in[1.2,2.0]$, and
$\epsilon\in[0.02,0.1]$ separately. The numerically obtained
$\beta_\mathrm{crit}$ are shown in Fig. \ref{bcrit_compare} in
comparison with Eq. \ref{iso_vsi_cond}, which generally agree. This
result, together with Fig. \ref{bcrit_compare1}, shows that  
$\beta \lesssim \beta_\mathrm{crit}$ can be used as a practical
criteria for the VSI to operate effectively in stably stratified,
vertically isothermal disks.

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.cm 0cm
  0cm]{figures/bcrit_compare_q.ps} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.0cm 0cm
  0.8cm]{figures/bcrit_compare_g.ps}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.0cm 0cm
  0.8cm]{figures/bcrit_compare_e.ps} 
  \caption{Dependence of the upper limit to the thermal relaxation timescale
    $\beta_\mathrm{crit}$ for the fundamental VSI mode on disk
    parameters. Our fiducial setup is $(\gamma, \Gamma= (1.4, 1.011)$
    and $(p,q,\epsilon)=(-1.5,-1,0.05)$. Top: varying 
    $q\in[-1.2,-0.2]$; middle: varying $\gamma\in[1.2,2.0]$; bottom:
    varying $\epsilon\in[0.02,0.1]$. The perturbation wavenumber is
    $\khat=10$.  
    \label{bcrit_compare}}  
\end{figure}

@@

% However, note that this issue is unimportant for low order modes 
% such as the fundamental mode, which are not sensitive to boundary
% conditions. This is useful because when the fundamental mode is
% stabilized as $\beta$ is increased, higher order modes will have
% already been stabilized, the the growth rate of surface modes also
% significantly reduced. Then as far as a characteristic upper limit to
% the thermal timescale is conconcered, we may just consider the
% fundamental mode.

Here we briefly consider a vertically non-isothermal disk model with  
$(p,q,\epsilon)=(0,-1,0.05)$ and $\Gamma=1.0837$. This value of 
$\Gamma$ is chosen so that the standard vertical domain size $\zmax =
5H$ effectively corresponds to the disk surface ($\zmax\simeq
H_s$). The disk temperature decreases rapidly with height such that 
$c_s^2(\zmax)\simeq 10^{-4}c_s^2(0)$.

In Fig. \ref{compare_modes_vnoniso_kx10} we plot the mode diagram for
$\khat=10$ for several values of $\beta$. This plot is 
qualitatively similar to Fig. \ref{compare_modes_cool_kx10}. As before
we find as $\beta$ is increased, higher order body modes are rapidly
stabilized, eventually only the fundamental mode can operate before
the VSI is entirely stabilized.

@@

We do not find surface modes for $\khat=10$. In the vertically
non-isothermal disk, surface modes develop for sufficiently large 
$\khat$ \citep{barker15}. We demonstrate them in
Fig. \ref{compare_modes_vnoniso_kx100} which show eigenvalues for
$\khat=100$. While growth rates generally decrease with increasing 
$\beta$, we find eigenvalues can overlap, making it difficult to 
distinguish between different modes. However, since there 
is a well-defined surface, we may seek the 
most unstable mode without the boundary-dependent issue that arose 
with a vertically isothermal disk.

@@

Fig. \ref{compare_eigenvz_vnoniso_kx100} show the most unstable modes
with $\khat=100$ and $\beta=10^{-3},\, 10^{-2}$ and
$0.1$. We find with $\beta=10^{-3}$ the most 
unstable mode is an asymmetric surface mode \citep[also seen in Fig. 7
of][]{barker15}. However, increasing the thermal timescale to
$\beta=10^{-2}$ an anti-symmetric surface mode becomes
dominant. Further increasing to $\beta=0.1$ surface modes are
suppressed and the dominant mode is the fundamental corrugation
mode. Notice as $\beta$ is increased the perturbation amplitude near
the vertical boundaries become negligible. This is not surprsing since
the stabilizing effect of buoyancy is largest at the disk surface.

@@

Finally, in Fig. \ref{bcrit_compare2} we plot the maximum VSI growth
rates in the vertically non-isothermal disk model.  Based on 
Eq. \ref{iso_vsi_cond}, we conjecture a critical
thermal timescale given by $\epsilon|s|/(\gamma - \Gamma)$. This
evaluates to 0.158, which is close to the cut-off thermal timescale
for $\khat\in[5,30]$. Fig. \ref{bcrit_compare2} is qualitatively quite
similar to the fiducial disk model in 
Fig. \ref{bcrit_compare1}. Importantly, the maximum thermal
timescale for $\khat=O(10)$ in fact provides a 
characteristic upper limit for other values of $\khat$.

>>>>

The modes shown in Fig. \ref{compare_modes_iso_kx10} are `body modes'
and are inertial waves destabilized by vertical shear
\citep{barker15}. (The exception being the fastest growing
mode for $\zmax=7H$ which is a surface mode --- see the discussion
below.)  The `fundamental corrugation mode', marked by `f',
corresponds to the eigenvalue with the smallest $|\sigma|$. The
fundamental mode is robust  
against vertical boundary conditions as it appears at the same
position in Fig. \ref{compare_modes_iso_kx10} for all cases. Higher  
order modes, corresponding to increasing $|\omega|$, increasingly
deviates from the analytic line due to the imposed boundaries, 
unless $\zmax$ is increased.

@@

\subsubsection{Comment on `surface modes'}\label{surf_comment} 
Fig. \ref{compare_modes_iso_kx10} lack the distinct `surface modes' found by 
\cite{nelson13}, \cite{mcnally14}  and \cite{barker15}. 
This a class of unstable modes associated with vertical shear near the
disk boundaries. They are absent in our  
analysis because there we assumed a disk with infinite 
vertical extent, since the vertically isothermal disk has no
surface. However, surface modes can occur in the present calculations 
because there is an imposed vertical boundary.

We illustrate surface modes in Fig. \ref{compare_modes_iso_kx30} 
which show unstable modes with $\khat=30$. Surface 
modes lie almost parallel to the vertical axis 
in such diagrams. They are readily identified for  $\zmax\geq 5H$ and
are marked by `s'.  Fig. \ref{lowfreq_eigenfunc_surf} shows an example
of a surface mode. We typically find an increasing
number of surface modes with increasing $\zmax$ and/or $\khat$. 
%consistent with \cite{barker15}.

@@

We find the number of unstable modes decrease with increasing
$\beta$. Consider the fiducial setup with $\zmax=5H$ 
(middle panel). The fundamental mode ($M=0$) is not the most unstable for   
$\beta=0.01$ as there area number of modes with $M>0$ that have larger
growth rates. As noted above, these higher order modes are affected by
$\zmax$, e.g. there is a significant mismatch between numerical and 
analytical results for $M\geq1$ when $\zmax=3H$ (upper panel).

@@

Fig. \ref{lowfreq_eigenfunc_cool}---\ref{lowfreq_eigenfunc_2d_cool}
shows the fundamental mode with $\khat=10$ in a disk with 
$\beta=0.1$. Comparing these plots with
Fig. \ref{lowfreq_eigenfunc}---\ref{lowfreq_eigenfunc_2d}, which have
$\beta=10^{-3}$, we see that increasing $\beta$ restricts the region 
characterized by $W\sim z$ and $\delta v_z\sim$ constant closer to the
midplane, with oscillatory behavior emerging near vertical boundaries
where buoyancy first becomes important. This leads to the `tilted'
pressure perturbations seen in Fig. \ref{lowfreq_eigenfunc_2d_cool} 
(cf. Fig. \ref{lowfreq_eigenfunc_2d}).

@@

Note that the curves are not neccessarily smooth because the most unstable
mode may not always correspond to the same type of disturbance. For example,
as seen above, for $\khat \leq 10$ higher order body modes are
dominant as $\beta\to 0$, but the fundamental 
mode becomes dominant as $\beta$ is increased. For $\khat=1$ the VSI
can operate slightly beyond $\beta_\mathrm{crit}$  (recall that
$\beta_\mathrm{crit}$ was derived assuming $\khat^2\gg 1$). For
$\khat=5,\,10$ (and $30$ for $\zmax=3H$), we find
$\beta_\mathrm{crit}$ accurately predicts the thermal timescal beyond
which the VSI is suppressed.

Complications from vertical boundaries arise when considering large
$\khat$. For $\khat=30$ and $\zmax=5H,\,7H$ we find for $\beta\gtrsim
0.1$ the dominant disturbance correspond to the leftmost eigenvalues 
seen in the lower two panels of Fig. \ref{compare_modes_cool_kx30},
which do not exist for $\zmax=3H$. For $\khat=50,\,100$ and
$\beta\to0$, we find surface modes dominate so the maximum growth rate
depends on $\zmax$ in the limit of rapid thermal relaxation. 
However, the maximum growth rate is insensitive to 
$\zmax$ for moderately large $\beta$ ($\gtrsim 0.1$), i.e. close to
the stabilization of the fundamental mode.  
%not surprising since we enable strongly stabilizing buoyancy forces
%at large zmax

@@

We further  test the dependence of $\beta_\mathrm{crit}$ on disk parameters
(Eq. \ref{iso_vsi_cond}).  We use the fiducial setup as a  reference
point and vary $q\in[-0.2,-1.2]$,  $\gamma\in[1.2,2.0]$, and
$\epsilon\in[0.02,0.1]$ separately. The numerically obtained
$\beta_\mathrm{crit}$ are shown in Fig. \ref{bcrit_compare} in
comparison with Eq. \ref{iso_vsi_cond}, which generally agree. This
result, together with Fig. \ref{bcrit_compare1}, shows that  
$\beta \lesssim \beta_\mathrm{crit}$ is a practical
criteria for the VSI to operate effectively in stably stratified,
vertically isothermal disks.

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.cm 0cm
  0cm]{figures/bcrit_compare_q.ps} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.0cm 0cm
  0.8cm]{figures/bcrit_compare_g.ps}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.0cm 0cm
  0.8cm]{figures/bcrit_compare_e.ps} 
  \caption{Dependence of the upper limit to the thermal relaxation timescale
    $\beta_\mathrm{crit}$ for the fundamental VSI mode on disk
    parameters. The fiducial setup is $(\gamma, \Gamma= (1.4, 1.011)$
    and $(p,q,\epsilon)=(-1.5,-1,0.05)$. Top: varying 
    $q\in[-1.2,-0.2]$; middle: varying $\gamma\in[1.2,2.0]$; bottom:
    varying $\epsilon\in[0.02,0.1]$. The perturbation wavenumber is
    $\khat=10$.  
    \label{bcrit_compare}}  
\end{figure}

@@

However, as we have seen in Fig. \ref{bcrit_compare1}, the maximum
growth rate becomes insensitive to $\zmax$ for $\beta\gtrsim
\beta_\mathrm{crit}$. This suggests that the characteristic upper
limit on the thermal timescale for the VSI is not sensitive to
$\zmax$, so that vertically isothermal disks are in fact suitable for
the purpose of this paper.

%Physically this is because modes that depend on vertical boundaries
%are also subject to the strongest stabilization by buoyancy, which
%maximize at $\zmax$. These modes are theremore always strongly damped
%as we increase the thermal timescale.

Nevertheless, here we briefly consider a vertically non-isothermal
disk model with $(p,q,\epsilon)=(0,-1,0.05)$ and $\Gamma=1.0837$. This
value of $\Gamma$ is chosen so that the standard vertical domain size $\zmax =
5H$ effectively corresponds to the disk surface ($\zmax\simeq
H_s$). The disk temperature decreases rapidly with height such that 
$c_s^2(\zmax)\simeq 10^{-4}c_s^2(0)$.

In Fig. \ref{compare_modes_vnoniso_kx10} we plot the mode diagram for
$\khat=10$ for several values of $\beta$. This plot is 
qualitatively similar to Fig. \ref{compare_modes_cool_kx10}. As before
we find as $\beta$ is increased, higher order body modes are rapidly
stabilized, eventually only the fundamental mode can operate before
the VSI is entirely suppressed.

@@

We do not find surface modes for $\khat=10$, but do find them for
sufficiently large $\khat$. We demonstrate this in  % In the vertically
% non-isothermal disk, surface modes develop for sufficiently large 
% $\khat$ \citep{barker15}. We demonstrate them in
Fig. \ref{compare_modes_vnoniso_kx100} which show eigenvalues for
$\khat=100$. While growth rates generally decrease with increasing 
$\beta$, we find eigenvalues can overlap, making it difficult to 
distinguish between different modes. However, since there 
is a well-defined surface, we may seek the  
most unstable mode without the boundary-dependent issue that arose 
with a vertically isothermal disk in the limit $\beta\to0$.

@@

Fig. \ref{compare_eigenvz_vnoniso_kx100} show the most unstable modes
with $\khat=100$ and $\beta=10^{-3},\, 10^{-2}$ and
$0.1$. For $\beta=10^{-3}$ the most 
unstable mode is an asymmetric surface mode \citep[also seen in Fig. 7
of][]{barker15}. However, increasing the thermal timescale to
$\beta=10^{-2}$ an anti-symmetric surface mode becomes
dominant. Further increasing to $\beta=0.1$, the dominant mode has a
mixed character between a surface mode and the fundamental
mode. Notice as $\beta$ is increased the perturbation amplitude near
the vertical boundaries become negligible. This is not surprsing since  
the stabilizing effect of buoyancy is largest at the disk surface. 
%formally diverges

@@

Finally, in Fig. \ref{bcrit_compare2} we plot the maximum VSI growth
rates in the vertically non-isothermal disk model. Based on 
Eq. \ref{iso_vsi_cond}, we conjecture the critical 
thermal timescale in the vertically non-isothermal disk is 
given by $\epsilon|s|/(\gamma - \Gamma)$. This 
evaluates to 0.158, which is close to the cut-off thermal timescale
for $\khat\in[5,30]$. Fig. \ref{bcrit_compare2} is qualitatively 
similar to the fiducial disk model in
Fig. \ref{bcrit_compare1}. Importantly, for $\khat=O(10)$ there is a
precise cut-off thermal timescale, the value of which in fact provides a  
characteristic upper limit for other $\khat$.
